---
title: "Final Project"
output: pdf_document
date: "2024-05-28"
---

```{r}
# Main idea is to identify how genes respond in two different conditions hypoxia and normxia
# Working with RNA-Seq data obtained from GEO
# Step 1: load the library and data set. Our data is in .tsv format

library(readr)
library(dplyr)
library(stringr)

raw_counts <- read_tsv("C:/Users/midhu/Downloads/GSE197576_raw_gene_counts_matrix.tsv.gz")
#checking the dimensions of data.
dim(raw_counts)
head(raw_counts)
#After checking the data we know that the data contains gene names and sample names
#Now we subset our data to specific samples required

Columns_to_select <- colnames(raw_counts)%>%
  stringr :: str_detect("sgCTRL|gene")

#Now we subset the data set based on the results above
counts_sub <- raw_counts[, Columns_to_select]
head(counts_sub)

#Now we convert the data into a matrix and remove the first column gene names
raw_counts_mat <- counts_sub[, -1]%>%
  as.matrix()
head(raw_counts_mat)

#Adding row names to the data set
rownames(raw_counts_mat) <- raw_counts$gene

head(raw_counts_mat)
head(counts_sub)

# Step 2: Calculate the total exon length per gene
# we need a package from bioconductor to identify the gene length 
library(TxDb.Hsapiens.UCSC.hg19.knownGene)

txdb <- TxDb.Hsapiens.UCSC.hg19.knownGene

genes(txdb)

#we just want the exons 
exons <- exonsBy(txdb, by = "gene")

#calculate the length of exons
width(exons)

#sum the exons width up per genes 
head(sum(width(exons)))

#Now let's make a tibble.we use enframe from tibble.The enframe() function is used to convert the result of the previous expression into a tibble with specified column names.
exon_len <- sum(width(exons))%>%
  tibble::enframe(name = "ENTREZID", value = "exon_length")

head(exon_len)

#Next, let’s map the ENTREZID to the official gene symbol so we can match the rownames of the RNAseq count matrix. We will need the org.Hs.eg.db Bioconductor package.This package contains info about gene,gene symbols,descriptions
library(org.Hs.eg.db)

map <- AnnotationDbi::select(org.Hs.eg.db, 
                            keys = exon_len$ENTREZID, 
                            columns= "SYMBOL",
                            keytype = "ENTREZID")

head(map)

#Now, we join the exon_len table 
map <- left_join(exon_len, map)

#checking the gene symbols present in raw_counts_mat and map
table(rownames(raw_counts_mat) %in% map$SYMBOL)

#Now we know the number of genes, so now lets see what genes are not in the mapping table we use setdiff to identify the genes present in raw_counts_mat

setdiff(rownames(raw_counts_mat), map$SYMBOL) %>%
  head(n = 20)

#subsetting the data
not_in_map <- setdiff(rownames(raw_counts_mat), map$SYMBOL)

raw_counts_mat[not_in_map, ]%>%
  head(n = 15)

#subset the common genes present in raw_count_mat and map file
common_genes <- intersect(rownames(raw_counts_mat), map$SYMBOL)

#select only the common genes and re-order them by common_genes
map <- map %>%
  dplyr::slice(match(common_genes, SYMBOL))

#we also subset the common genes and re-order them in raw_count_mat
raw_counts_mat <- raw_counts_mat[common_genes, ]
head(raw_counts_mat)

#Now the order of gene is same for map and raw_counts_mat

``


#Step 3: Normalizing the raw counts to TPM
```{r}
#creating a TPM function to normalize the data
count2tpm <- function(count_matrix, exon_length){
  reads_per_bp_gene <- count_matrix/ exon_length
  reads_per_bp_sample <- colSums(reads_per_bp_gene)
  tpm_matrix <- t(t(reads_per_bp_gene)/reads_per_bp_sample) * 1000000
  return(tpm_matrix)
}

#Now lets apply the tpm function to required data
tpm <- count2tpm(raw_counts_mat, map$exon_length)
head(tpm)

#Analyzing the gene expression values using t-test
#First we'll start with WASH7P gene. 
t.test(tpm["WASH7P", c(1,2)], tpm["WASH7P", c(3,4)])
#we see there's no significance difference as the p-value is 0.3

#Second gene VEGFA
t.test(tpm["VEGFA", c(1,2)], tpm["VEGFA", c(3,4)])
#We see that there's a significant difference between the 2 conditions in VEGFA gene

#Third Gene SLC2A1 
t.test(tpm["SLC2A1", c(1,2)], tpm["SLC2A1", c(3,4)])
# The p-value is 0.01354, indicating a significant difference in SLC2A1 expression between the two conditions

#Now We analyze all genes in our data set by creating a custom function 
mytest <- function(x) t.test(x[c(1,2)], x[c(3,4)], var.equal = TRUE)$p.value
pvals <- apply(tpm, 1, mytest)%>%
  na.omit()

head(pvals)

#Analyzing the gene expression data with ggplot
library(ggplot2)
library(tidyr)
library(dplyr)

#Loading the p-values
pval_df <- pvals %>%
  tibble::enframe(name = "gene", value = "pvalue")

head(pval_df)

#Creating a histogram to visualize the distribution of p-values
ggplot(pval_df, aes(x = pvalue)) +
  geom_histogram(color = "white") +
  theme_bw(base_size = 14) +
  ggtitle("p-value distribution")

#Now, we calculate the average expression levels for both conditions and identifying the up-regulated genes.
avg_normoxia <- rowMeans(tpm[, c(1, 2)])
avg_hypoxia <- rowMeans(tpm[,c(3,4)])

#We identify the up regulated genes
up_genes <- (avg_hypoxia - avg_normoxia) > 0 
up_gene_names <- rownames(tpm)[up_genes]
head(up_gene_names)

#we identified the genes that are upregulated, now we select the differentially expressed genes
differential_genes <- pvals[pvals < 0.01 & !is.na(pvals)] %>%
  names()
differential_up_genes <- intersect(differential_genes, up_gene_names)

#checking the count
length(differential_up_genes)

#We prepare our data for creating a boxplot to visualize gene expression changes between the two conditions. 
#Converting gene expression data to long format
tpm[differential_genes, ] %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "gene") %>%
  tidyr::pivot_longer(-1, names_to = "sample", values_to = "tpm")

#We add one more column to denote the condition 
tpm_df<- tpm[differential_genes, ] %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var="gene") %>%
  tidyr::pivot_longer(-1, names_to = "sample", values_to = "tpm") %>%
  tidyr::separate(sample, into = c("sample", "condition"), sep = "_sgCTRL_")

head(tpm_df)

#creating a boxplot
ggplot(tpm_df, aes(x = condition, y = log2(tpm + 1))) +
  geom_boxplot() +
  theme_bw(base_size = 14)

#Visualizing raw expression values to identify outliers. we select specific genes to check this 
ggplot(tpm_df %>%
         filter(gene %in% c("VEGFA", "SLC2A1")), 
       aes(x = condition, y = log2(tpm + 1))) +
  geom_point() +
  geom_boxplot() +
  facet_wrap(~ gene) +
  theme_bw(base_size = 14)

## FDR 
#order the pvals computed above and plot it.
alpha <- 0.05

#m is the number of 17744 comparisons 
m <- length(pvals)

# let's arrange the p-value from small to big and get only the first 5000  
top_5000_pvalue <- pval_df %>%
  dplyr::arrange(pvalue) %>%
  mutate(rank = row_number()) %>%
  dplyr::slice(1:5000)

head(top_5000_pvalue)

```

## Analyzing gene expression data with DESeq2
```{r}
#Loading required libraries
library(DESeq2)
coldata <- data.frame(condition = c("normoxia", "normoxia", "hypoxia", "hypoxia"))
rownames(coldata) <- colnames(raw_counts_mat)

coldata

#creating a DESeqData set object 
dds <- DESeqDataSetFromMatrix(countData = raw_counts_mat,
                              colData = coldata,
                              design = ~ condition)

#Performing differential gene expression analysis
dds <- DESeq(dds)

#we extract results from the DESeq2 analysis, specifying the contrast between two conditions
res <- results(dds, contrast = c("condition", "hypoxia", "normoxia"))%>%
  na.omit()

#Generating a volcano plot using ggplot
ggplot(data = as.data.frame(res))+
  geom_point(aes(x = log2FoldChange, y = -log10(pvalue)))+
  ggtitle("Volcano plot")
#Explanation: In this plot,we can see that each point represents a gene. The x-axis shows how much a gene's expression changes (log2 fold change), while the y-axis indicates the significance of the change (-log10 p-value). Genes with a significant change will be far to the left or right on the plot, and those with very low p-values will be at the top.

#Examining the top differentially expressed genes based on padj and log2FoldChange
top_differentially_expressed_genes <- res %>%
  as.data.frame() %>%
  arrange(padj, desc(log2FoldChange)) %>%
  head(n = 30)

#We are trying to label the genes in Volcano plot here
genes_to_label <- res %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "gene") %>%
  filter(!stringr::str_detect(gene, "LOC"),
         abs(log2FoldChange) >= 2.5,
         padj <= 0.001)

#Loading library required for labeling
library(ggrepel)

#Generating volcano plot with labels using ggrepel
ggplot(data = as.data.frame(res), aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point() +
  ggrepel::geom_label_repel(data = genes_to_label, aes(label = gene)) +
  theme_bw(base_size = 14)

#To further enhance the plot, we can color the points based on significance
res2 <- res %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "gene") %>%
  mutate(sig = case_when(
    !stringr::str_detect(gene, "LOC") &
      abs(log2FoldChange) >= 2.5 &
      padj <= 0.001 ~ "sig",
    TRUE ~ "not sig"
  ))

#Now we update the volcano plot with signficant and not significant genes
ggplot(res2, aes(x = log2FoldChange, y = -log10(pvalue))) +
  geom_point(aes(color = sig)) +
  ggrepel::geom_label_repel(data = genes_to_label, aes(label = gene))+
  theme_bw(base_size = 14)

```

# PCA Analysis
```{r}
#Performing variance stabilizing transformation on the data
vsd <- vst(dds, blind = FALSE)

#Plot PCA using vst data
plotPCA(vsd, intgroup=c("condition"))

#Filtering significant genes based on adjusted p-value and log2 fold change for plotting heatmap
library(ComplexHeatmap)
significant_genes <- res %>%
  as.data.frame()%>%
  filter(padj <= 0.01, abs(log2FoldChange)>= 2)%>%
  rownames()

#To ensure the heatmap is informative, we scale the data using the scale function. It sets mean to 0 and std dev to 1
normalized_counts <- assay(vsd) %>% as.matrix()
significant_mat <- normalized_counts[significant_genes, ]

#Generating heatmap with scaled data
Heatmap(
  t(scale(t(significant_mat))),
  show_row_names = FALSE,
  name = "scaled\nexpression"
)

#creating a data frame for adding annotations to heatmap. This helps us understand the heatmap a little better
coldata <- data.frame(
  condition = c("hypoxia", "hypoxia", "normoxia", "normoxia")
  )

#Creating a HeatmapAnnotation object for the column annotations
col_anno <- HeatmapAnnotation(
  df = coldata,
  col = list(condition = c("hypoxia" = "green", "normoxia" = "blue"))
)

#Generating heatmap with column annotations
Heatmap(
  t(scale(t(significant_mat))),
  top_annotation = col_anno,
  show_row_names = FALSE,
  name = "scaled normalized\nexpression"
)

```


```{r}
## Pathway Analysis
library(clusterProfiler)

#Converting gene ID to EntrezId
significant_genes_map<- clusterProfiler::bitr(geneID = significant_genes,
                      fromType="SYMBOL", toType="ENTREZID",
                      OrgDb="org.Hs.eg.db")

head(significant_genes_map)

# we also require backrground genes.Background genes are genes that are detected in the RNAseq experiment. we are also filtering out the genes that has zero expression in all conditions i.e those are not detected in the RNA-sequencing


background_genes <- res %>% 
  as.data.frame() %>% 
  filter(baseMean != 0) %>%
  tibble::rownames_to_column(var = "gene") %>%
  pull(gene)

head(background_genes)
#As we are gonna use this results often saved it to a dataframe for easy accesbility.
 res_df <- res %>% 
  as.data.frame() %>% 
  filter(baseMean != 0) %>%
  tibble::rownames_to_column(var = "gene")

#Now, we also the map the gene symbols and their corresponding Entrez IDs for the background genes. 
background_genes_map <- clusterProfiler::bitr(geneID = background_genes,
                                               fromType = "SYMBOL",
                                               toType = "ENTREZID",
                                               OrgDb = "org.Hs.eg.db")
 
#Now, we perform gene enrich analysis using enrichGO

ego <- enrichGO(
  gene = significant_genes_map$ENTREZID,
  universe = background_genes_map$ENTREZID,
  OrgDb = org.Hs.eg.db,
  keyType = "ENTREZID",
  ont = "BP",
  pvalueCutoff = 0.01,
  pAdjustMethod = "BH",
  qvalueCutoff = 0.05,
  readable = TRUE,
)
head(ego)
#Now let's do a barplot for visualization
barplot(ego, showCategory=20) 
dotplot(ego)

## Gene Enrichment Analysis. MSigDBis a widely used resource for gene set enrichment analysis.
library(msigdbr)
#We retrieve Hallmark gene sets for Homo sapiens
m_df <- msigdbr(species = "Homo sapiens")

#selecting the gene set name and ENTREZ ID
m_t2g <- msigdbr(species = "Homo sapiens", category = "H") %>% 
  dplyr::select(gs_name, entrez_gene)

head(m_t2g)

#Now, we'll use siginifcant gene set and Hallmart gene set to perform gene enrichment analysis
em <- enricher(significant_genes_map$ENTREZID, TERM2GENE = m_t2g, universe = background_genes_map$ENTREZID)

head(em)

dotplot(em)

#From the graph we can say that "HALLMARK HYPOXIA" is the pathway most strongly associated with the genes and that is why we are studying it.


## Gene set enrichment analysis: Gene Set Enrichment Analysis (GSEA) is an alternative approach that considers the entire list of genes ranked by their association with a phenotype or condition. 
res_df <- res_df %>% 
  mutate(signed_rank_stats = sign(log2FoldChange) * -log10(pvalue)) %>%
  left_join(background_genes_map, by = c("gene" = "SYMBOL")) %>%
  arrange(desc(signed_rank_stats))

#As the infinite values are present in the data we replace the infinite values with large numbers

res_df <- res_df %>%
  mutate(negative_log10pvalue = -log10(pvalue)) %>%
  mutate(negative_log10pvalue = ifelse(is.infinite(negative_log10pvalue), 1000, negative_log10pvalue)) %>%
  mutate(signed_rank_stats = sign(log2FoldChange) * negative_log10pvalue)

#Now, we perform gsea using pre-ranked gene list and a reference gene set(m_t2g)
gene_list <- res_df$signed_rank_stats
names(gene_list) <- res_df$ENTREZID

em2 <- GSEA(gene_list, TERM2GENE = m_t2g)

head(em2)

#ES measures the degree ton whcich the gene set is overrepresented. We can see for hypoxia condition they are positively upregulated 

#viewing results 
em2@result

#visualizing the results using gseaplot function

p1<- gseaplot(em2, geneSetID = "HALLMARK_G2M_CHECKPOINT", 
              by = "runningScore", title = "HALLMARK_G2M_CHECKPOINT")
p2 <- gseaplot(em2,geneSetID = "HALLMARK_HYPOXIA",
               by = "runningScore", title = "HALLMARKHYPOXIA")

#Comparing the visulaization side by side
p1/p2


#From the graph we can see that hypoxia gene sets are at the front of the ranked list which means hypoxia pathway is upregulated and the G2M gene sets are at the end of the ranked list which means it is downregulated

```


